The Close Companion Mass-Ratio Distribution of Intermediate-Mass Stars

### 2016-04-19 paper/ms.tex
Version after resubmission to journal.

\begin{abstract}
Binary stars and higher-order multiple systems are a ubiquitous outcome of star formation, especially as the system mass increases. The companion mass-ratio distribution is a unique probe into the conditions of the collapsing cloud core and circumstellar disk(s) of the binary fragments. Inside $a \sim 1000$ AU the disks from the two forming stars can interact, and additionally companions can form directly through disk fragmentation. We should therefore expect the mass-ratio distribution of close companions ($a \lesssim 100$ AU) to differ from that of wide companions. This prediction is difficult to test with intermediate-mass primary stars using traditional methods because the contrast ratios that would be required to detect low-mass companions at narrow working angles are not yet achievable. We present the results of a survey searching for companions to A- and B-type stars using the direct spectral detection technique, which is sensitive to late-type companions within $\sim 1''$ of the primary and which has no inner working angle. We estimate the temperatures and surface gravity of most of the 341 sample stars, and derive their masses and ages. We additionally estimate the temperatures and masses of the 64 companions we find, 23 of which are new detections. We find that the mass-ratio distribution for our sample has a maximum near $q \sim 0.3$. Our mass-ratio distribution has a very different form than in previous work, where it is usually well-described by a power law, and indicates that close companions to intermediate-mass stars experience significantly different accretion histories or formation mechanisms than wide companions.
\end{abstract}
\keywords{binaries: spectroscopic, stars:early-type, stars:formation, stars:statistics}

@@

The overall multiplicity rate and the distributions of mass ratio, period, and eccentricity of a binary star population place important constraints on the mode of binary star formation. While the period and eccentricity are altered by dynamical processing in the birth cluster, the present-day mass ratio of a binary system is a direct result of its formation \citep{Parker2013}. Most binary stars are thought to form via core fragmentation \citep{Boss1979, Boss1986, Bate1995}, in which a collapsing core fragments into two or more individual protostars. The number and initial masses of the fragments are set by the total core mass, as well as its rotation, turbulence, and its temperature and density structure. If the fragments are well separated ($a \gtrsim 1000$ AU), they will evolve independently of each other, accreting mass from the core material onto their own protostellar disks and then onto the protostars themselves. However close fragments ($a \sim 100$ AU) will interact with each other; the protostellar disk may be truncated, destabilized, or form into a circumbinary disk if the separation is small enough \citep{Bate1997}. In addition, an unstable disk can fragment to form low-mass companions \citep{Kratter2006, Stamatellos2011}. The mass ratios of close companions formed via either mechanism should be affected by preferential accretion. Most work has suggested that the disk material will preferentially accrete onto the lower mass companion \citep{Bate1997, BBB2002}; however, recent work has indicated that magnetic disk braking may result in preferential accretion onto the more massive component \citep{Zhao2013} instead. In either case, we would expect to find a mass-ratio distribution for companions inside a few $100$ AU that differs from that of companions on wider orbits.

The mass ratio, period, and eccentricity distributions are well-known for solar type stars \citep{Duquennoy1991, Raghavan2010} and cooler stars \citep{Fischer1992, Delfosse2004}. Interestingly, \citet{Reggiani2011}, and later \citet{Meyer2013}, found that the mass-ratio distribution of field solar-type and M-dwarf stars is invariant to separation. However, the solar-type binary semimajor axis distribution peaks near $45$ AU, and only $\sim 10\%$ of binary systems have separations $a < 100$ AU \citep{Raghavan2010}; the 30 stars used in the analysis by \citet{Reggiani2011} is unsufficient to rule out a separation-variant mass-ratio distribution if it changes outside $a \gtrsim 100$ AU. Field M-dwarf binary systems have typical separations of $\sim 5$ AU \citep{Fischer1992}, and so the 27 stars using in the analysis is again insufficient to look for differences outside of $100$ AU.

All of the orbital distributions are much less certain for more massive stars. The reason for this is two-fold: first, more massive stars tend to be more distant than sunlike or low-mass stars, meaning many of the companions are angularly close to the primaries and difficult to detect with imaging techniques. Second, the primary stars tend to be rapid rotators, which limits radial velocity precision to $\sim 1 \mathrm{km\ s}^{-1}$. Additionally, radial velocity monitoring can only measure a mass ratio if spectral lines from both components are visible; this typically suffers from the same flux ratio difficulty as imaging techniques.

@@

The sample, being comprised of early-type stars, is heavily biased towards young stars. The estimated ages range from about 10 Myr to 1 Gyr, with most falling in the range of a few tens or hundreds of Myrs. The sample also mostly contains nearby stars, although the magnitude limit provides a greater extent than a volume-limited survey would have. The parallactic distances in our sample range from $\sim 15 - 2000$ pc. The maximum detectable separation, which we define as the point at which the companion is no longer guaranteed to fall on the spectrograph slit (see Section \ref{subsec:specdata} for a description of the spectrographs we use), is $\sim 20 - 4000$ AU. The median parallactic distance in our sample is $95$ pc, corresponding to a projected separation of $\sim 200$ AU. Most of the companions we are able to detect are close enough to have been impacted by the circumprimary disk, with $85\%$ of the sample sensitive to companions inside $100$ AU.

@@

\subsection{Imaging Data}
As part of the follow-up effort, we used the NIRI instrument behind the Altair adaptive optics system on the Gemini North Telescope. For each star listed in Table \ref{tab:imaging_obs}, we obtained 25 images in 5 dithering positions. We used the K-continuum band centered on $2.2718\ \mu m$ and a variety of exposure times and dates (listed in Table \ref{tab:imaging_obs}). Because the targets are all extremely bright, we used the high read noise and high flux detector settings to allow for very short co-add exposure times. We reduced the data using the Gemini set of IRAF tasks, which include steps for nonlinearity correction, flat-fielding, sky subtraction, and co-addition of the dither frames.

@@

It is also possible that some of the companions found in this work were formed from a gravitationally unstable disk \citep[e.g.][]{Kratter2006, Stamatellos2011}. Being a completely different formation mechanism than the way wide companions form, we would expect the initial companion mass function to differ. Additionally, such companions would undergo the same preferential accretion discussed above. This explanation may even be preferable, since it could explain why we see a separation-dependent mass-ratio distribution around intermediate-mass stars but not low-mass stars: more massive stars tend to have more massive disks \citep{Andrews2013} that are more likely to fragment \citep{Kratter2010}.

@@

\section*{Acknowledgements}
This research has made use of the SIMBAD database, operated at CDS, Strasbourg, France, and of Astropy, a community-developed core Python package for Astronomy (Astropy Collaboration, 2013).
It was supported by a start-up grant to Adam Kraus as well as a University of Texas Continuing Fellowship and a Dissertation Writing Fellowship to Kevin Gullikson.

>>>>

\begin{abstract}
Binary stars and higher-order multiple systems are an ubiquitous outcome of star formation, especially as the system mass increases. The companion mass-ratio distribution is a unique probe into the conditions of the collapsing cloud core and circumstellar disk(s) of the binary fragments. Inside $a \sim 1000$ AU the disks from the two forming stars can interact, and additionally companions can form directly through disk fragmentation. We should therefore expect the mass-ratio distribution of close companions ($a \lesssim 100$ AU) to differ from that of wide companions. This prediction is difficult to test using traditional methods, especially with intermediate-mass primary stars, for a variety of observational reasons. We present the results of a survey searching for companions to A- and B-type stars using the direct spectral detection method, which is sensitive to late-type companions within $\sim 1''$ of the primary and which has no inner working angle. We estimate the temperatures and surface gravity of most of the 341 sample stars, and derive their masses and ages. We additionally estimate the temperatures and masses of the 64 companions we find, 23 of which are new detections. We find that the mass-ratio distribution for our sample has a maximum near $q \sim 0.3$. Our mass-ratio distribution has a very different form than in previous work, where it is usually well-described by a power law, and indicates that close companions to intermediate-mass stars experience significantly different accretion histories or formation mechanisms than wide companions.
\end{abstract}
\keywords{binaries: spectroscopic, stars:early-type, stars:formation, stars:statistics}

@@

The overall multiplicity rate and the distributions of mass ratio, period, and eccentricity of a binary star population place important constraints on the mode of binary star formation. While the period and eccentricity are altered by dynamical processing in the birth cluster, the present-day mass ratio of a binary system is a direct result of its formation \citep{Parker2013}. Most binary stars are thought to form via core fragmentation \citep{Boss1979, Boss1986, Bate1995}, in which a collapsing core fragments into two or more individual protostars. The number and initial masses of the fragments are set by the total core mass, as well as its rotation, turbulence, and its temperature and density structure. If the fragments are well separated ($a \gtrsim 1000$ AU), they will evolve independently of each other, accreting mass from the core material onto their own protostellar disks and then onto the protostars themselves. However close fragments ($a \sim 100$ AU) will interact with each other; the protostellar disk may be truncated, destabilized, or form into a circumbinary disk if the separation is small enough \citep{Bate1997}. In addition, an unstable disk can fragment to form low-mass companions \citep{Kratter2006, Stamatellos2011}. The mass ratios of close companions formed via either mechanism should be affected by preferential accretion. Most work has suggested that the disk material will preferentially accrete onto the lower mass companion \citep{Bate1997, BBB2002}; however, recent work has indicated that magnetic disk braking may result in preferential accretion onto the more massive component \citep{Zhao2013} instead. In either case, we would expect to find a mass-ratio distribution for companions inside a few $100$ AU that differs from that of companions on wider orbits where preferential accretion does not occur.

The mass ratio, period, and eccentricity distributions are well-known for solar type stars \citep{Duquennoy1991, Raghavan2010} and cooler stars \citep{Fischer1992, Delfosse2004}. Interestingly, \citet{Reggiani2011}, and later \citet{Meyer2013}, found that the mass-ratio distribution of field solar-type and M-dwarf stars is invariant to separation. The field M-dwarf semimajor axis distribution peaks near $\sim 5$ AU \citep{Duchene2013}, with very few companions at separation $\gtrsim 100$ AU; the 27 stars used in the analysis by \citet{Reggiani2011} is unsufficient to compare the mass-ratio distribution inside $\sim 100$ AU with that outside it. However, the solar-type period distribution peaks near $45$ AU \citep{Raghavan2010}, with roughly $40\%$ of binary systems on orbits wider than $100$ AU. The nondetection of a difference in mass-ratio distribution is significant, although with only 30 stars in the field sample it is difficult to completely rule out that such a difference exists.

All of the orbital distributions are much less certain for more massive stars. The reason for this is two-fold: first, more massive stars tend to be more distant than sunlike or low-mass stars, meaning many of the companions are angularly close to the primaries and difficult to detect with imaging techniques. Second, the primary stars tend to be rapid rotators, which limits radial velocity precision to $\sim 1 \mathrm{km\ s}^{-1}$ and causes the spectral lines of double-lined systems to blend. Radial velocity monitoring can only measure a mass ratio if spectral lines from both components are visible and separable; this typically suffers from the same flux ratio difficulty as imaging techniques.

@@

The sample, being comprised of early-type stars, is heavily biased towards young stars. The estimated ages range from about 10 Myr to 1 Gyr, with most falling in the range of a few tens or hundreds of Myrs. The sample also mostly contains nearby stars, although the magnitude limit provides a greater extent than a volume-limited survey would. The parallactic distances in our sample range from $\sim 15 - 2000$ pc. The maximum detectable separation, which we define as the point at which the companion is no longer guaranteed to fall on the spectrograph slit (see Section \ref{subsec:specdata} for a description of the spectrographs we use), is $\sim 20 - 4000$ AU. The median parallactic distance in our sample is $95$ pc, corresponding to a projected separation of $\sim 200$ AU. Most of the companions we are able to detect are close enough to have been impacted by the circumprimary disk, with $85\%$ of the sample sensitive to companions inside $100$ AU.

@@

\subsection{Imaging Data}
As part of the follow-up effort, we used the NIRI instrument behind the Altair adaptive optics system on the Gemini North Telescope. For each star listed in Table \ref{tab:imaging_obs}, we obtained 25 images in 5 dithering positions. We used the K-continuum band centered on $2.2718\ \mu m$ and a variety of exposure times and dates (listed in Table \ref{tab:imaging_obs}). Because the targets are all extremely bright, we used the high read noise and high flux detector settings to allow for very short co-add exposure times and to prevent saturation. We reduced the data using the Gemini set of IRAF tasks, which include steps for nonlinearity correction, flat-fielding, sky subtraction, and co-addition of the dither frames.

@@

It is also possible that some of the companions found in this work were formed from a gravitationally unstable disk \citep[e.g.][]{Kratter2006, Stamatellos2011}. Being a completely different formation mechanism than the way wide companions form, we would expect the initial companion mass function to differ. Such companions would undergo the same preferential accretion discussed above.

@@

\section*{Acknowledgements}
We would like to thank the anonymous referee for their helpful comments, which have notably improved this work. This research has made use of the SIMBAD database, operated at CDS, Strasbourg, France, and of Astropy, a community-developed core Python package for Astronomy (Astropy Collaboration, 2013).
It was supported by a start-up grant to Adam Kraus as well as a University of Texas Continuing Fellowship and a Dissertation Writing Fellowship to Kevin Gullikson.
